taxadb: A High-Performance Local Taxonomic Database Interface

### 2020-02-08 paper/manuscript.Rmd
Change IUCN example data column name to consistently refer to IUCN by name rather than by "redlist".

printtable <- function(df, ...){
  df %>%
    kableExtra::kable("latex", booktabs=T, ...) %>%
    kableExtra::kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)

@@

While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015],
and a growing number of databases and tools have emerged and grown over the past few decades
[e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018],
it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files which are 
automatically downloaded and imported and stored on a local database by `taxadb`.  
The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes
over a GB, which can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection.  
By using a local database as the backend, `taxadb` allows R users to interact with large data files without large memory (RAM) requirements.  A query for a single name over the web API requires a remote server to respond, execute the query, 
and serialize the response, which can take several seconds. Thus it does not take many taxa before transferring the 
entire data set to query locally is more efficient.  Moreover, this local copy can be cached on the user's machine, 
requiring only the one-time setup, and enabling offline use and reproducible queries.  Rather than returning data
in whatever format is given by the provider, `taxadb` provides a data structure following a consistent,
standardized layout or schema following the Darwin Core Standard [@Wieczorek2012].  Table 1 summarizes the list of all naming providers currently accessed by `taxadb`.  More details are provided in the Data Sources Vignette, <https://docs.ropensci.org/taxadb/articles/data-sources.html>.

@@

tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
                                 "National Center for Biological Information's Taxonomy database (NCBI)",
                                 "Catalogue of Life (COL)",
                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
                                 "FishBase (Froese and Pauly 2019)",
                                 "Open Tree Taxonomy (OTT)",
                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE, 
               caption = "Descriptions of providers supported by taxadb.") %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")

@@

After loading the package, we look up the taxonomic identifier for a Atlantic Cod, *Gadus morhua*, and the compliment:

@@

Our first call to any `taxadb` functions will automatically set up a local, persistent database if one has not yet been created. This one-time setup will download, extract, and import the compressed data into persistent database storage (using the appropriate location specified by the operating system [see @rappdirs], or configured using the environmental variable `TAXADB_HOME`).  The example above searches for names in ITIS, the default provider, which can be configured using the `provider` argument. Any future function calls to this function or any other function using data from the same names provider will be able to access this data rapidly without the need for processing or an internet connection.

Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 6), or simply using `all` to install all available providers:

@@

`taxadb` functions like `get_ids()` and `td_create()` take an optional argument, `db`, to an external database connection.  `taxadb` will work with most DBI-compliant databases such as MySQL or Postgres, but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.  These latter options are also much easier for most users, since each can be installed directly as an R package. `taxadb` will default to the fastest available option.  `taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the large names providers.

`taxadb` uses the widely known SQLite database by default, but users are encouraged to install the optional, suggested database backends by passing the option `dependencies = TRUE` to the install command.  This installs a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage.  This also installs `duckdb`, another local columnar database which is rapidly emerging as an alternative to MonetDB and SQLite. `taxadb` will automatically detect and use these database engines if available, and automatically handles opening, caching, and closing the database connection. For large queries, MonetDBLite or `duckdb` deliver impressive improvements.  Our benchmark on resolving the 750 species names in in the Breeding Bird Survey against over 3 million names known in the 2019 Catalogue of Life takes 8 minutes in SQLite but less than a second in MonetDBLite.

@@

These issues are best illustrated using a minimal example.  We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species (Table 1, 2).

@@

```{r iucn_table, echo=FALSE, cache = FALSE}
status_data %>% printtable(caption = "IUCN status for the selected 10 species.")
```

```{r trait_table, echo = FALSE, cache = FALSE }
trait_data %>% printtable(caption = "Trait data for the selected 10 species.")
```

If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved (Table 3).

```{r}
joined <- full_join(trait_data, status_data, by = c("elton_name" = "redlist_name")) 
```

```{r echo = FALSE, cache = FALSE}
joined %>%
  printtable(caption = "IUCN and trait data joined on scientific name.")
```

@@

traits <- trait_data %>% mutate(id = get_ids(elton_name, "col"))
status <- status_data %>% mutate(id = get_ids(redlist_name, "col"))
```

@@

```{r cache = FALSE, echo = FALSE}
## Just for pretty-printing
joined %>%  
  tidyr::replace_na(list(category = "-", elton_name = "-", redlist_name = "-")) %>%
  select(elton_name, redlist_name, mass, category, id) %>%
  printtable(caption = "IUCN and trait data joined on taxonomic ID.")
```

This results in many more matches (Table 4), as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.  While we have focused on a small example for visual clarity here, the `get_ids()` function in `taxadb` can quickly resolve hundreds of thousands of species names to unique identifiers, thanks to the performance of fast joins in a local MonetDBLite database.

@@

```{r}
multi_match <- filter_name("Zosterops rendovae")
```

```{r echo = FALSE, cache = FALSE}
multi_match %>%
  select(1:5) %>%
  printtable(caption = "Accepted name matches for Zosterops rendovae.")
```

We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between (Table 5).
This is an example of how `taxadb` seeks to provide users with information from existing authorities and names providers,
rather than make a potentially arbitrary decision.  Because they return `data.frame`s,  `filter_*` functions provide both potential matches.  Note that the simpler `get_` functions (`get_ids()`) consider multiple name as `NA` for the `id`, making them suitable for automated pipelines where manual resolution of duplicates is not an option.

@@

The full taxonomic record in the database can also be directly accessed by `taxa_tbl()`, allowing for whole-database queries that are not possible through the API or web interface of many providers.  For example, we can easily check the coverage of accepted species names in each of the classes of vertabrates within the Catalogue of Life:

>>>>

printtable <- function(df, ...){
  df %>%
    kableExtra::kable("latex", booktabs=T, ...) %>%
    kableExtra::kable_styling(full_width = F, latex_options = "hold_position")
}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)

@@

While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015],
and a growing number of databases and tools have emerged over the past few decades
[e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018],
it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files which are 
automatically downloaded, imported, and stored on a local database by `taxadb`. 
The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes
over a GB, which can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection. 
By using a local database as the backend, `taxadb` allows R users to interact with large data files without large memory (RAM) requirements.  A query for a single name over the web API requires a remote server to respond, execute the query, 
and serialize the response, which can take several seconds. Thus it does not take many taxa before transferring the 
entire data set to query locally is more efficient.  Moreover, this local copy can be cached on the user's machine, 
requiring only the one-time setup, and enabling offline use and reproducible queries.  Rather than returning data
in whatever format is given by the provider, `taxadb` provides a data structure following a consistent,
standardized layout or schema following the Darwin Core Standard [@Wieczorek2012].  Table 1 summarizes the list of all naming providers currently accessed by `taxadb`.  More details are provided in the Data Sources Vignette, <https://docs.ropensci.org/taxadb/articles/data-sources.html>.

@@

tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
                                 "National Center for Biological Information's Taxonomy database (NCBI)",
                                 "Catalogue of Life (COL)",
                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
                                 "FishBase (Froese and Pauly 2019)",
                                 "Open Tree Taxonomy (OTT)",
                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE, 
               caption = "Descriptions of the providers supported by taxadb with their reference abbreviation and the total number of identifiers contained by each provider.") %>%
  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")

@@

After loading the package, we look up the taxonomic identifier for Atlantic Cod, *Gadus morhua*, and the compliment:

@@

Our first call to any `taxadb` functions will automatically set up a local, persistent database if one has not yet been created. This one-time setup will download, extract, and import the compressed data into persistent database storage (using the appropriate location specified by the operating system [see @rappdirs], or configured using the environmental variable `TAXADB_HOME`).  The example above searches for names in ITIS, the default provider, which can be configured using the `provider` argument. Any future function calls to this function or any other function using data from the same provider will be able to access this data rapidly without the need for processing or an internet connection.

Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 1), or simply using `all` to install all available providers:

@@

`taxadb` functions like `get_ids()` and `td_create()` take an optional argument, `db`, to an external database connection.  `taxadb` will work with most DBI-compliant databases such as MySQL or Postgres, but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.  These latter options are also much easier for most users, since each can be installed directly as an R package. `taxadb` will default to the fastest available option.  `taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the larger providers.

`taxadb` uses the widely known SQLite database by default, but users are encouraged to install the optional, suggested database backends by passing the option `dependencies = TRUE` to the install command.  This installs a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage.  This also installs `duckdb`, another local columnar database which is rapidly emerging as an alternative to MonetDB and SQLite. `taxadb` will automatically detect and use these database engines if available, and automatically handles opening, caching, and closing the database connection. For large queries, MonetDBLite or `duckdb` deliver impressive improvements.  Our benchmark on resolving the 750 species names in the Breeding Bird Survey against over 3 million names known in the 2019 Catalogue of Life takes 8 minutes in SQLite but less than a second in MonetDBLite.

@@

These issues are best illustrated using a minimal example.  We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits].  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species (Table 2, 3).

@@

```{r iucn_table, echo=FALSE, cache = FALSE}
status_data %>% printtable(caption = "The subset of the IUCN status data used for subsequent taxonomic identifier examples.") 
```

```{r trait_table, echo = FALSE, cache = FALSE }
trait_data %>% printtable(caption = "The subset of the Elton trait data used for subsequent taxonomic identifier examples.")
```

If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved (Table 4).

```{r}
joined <- full_join(trait_data, status_data, by = c("elton_name" = "iucn_name")) 
```

```{r echo = FALSE, cache = FALSE}
joined %>%
  printtable(caption = "Example IUCN and trait data joined directly on scientific name showing only one match. While common, joining on scientific name does not account for taxnomic inconsistiencies between databases and therefore results in seemingly very little overlap in species representation between the two.")
```

@@

traits <- trait_data %>% mutate(id = get_ids(elton_name, "col"))
status <- status_data %>% mutate(id = get_ids(iucn_name, "col"))
```

@@

```{r cache = FALSE, echo = FALSE}
## Just for pretty-printing
joined %>%  
  tidyr::replace_na(list(category = "-", elton_name = "-", iucn_name = "-")) %>%
  select(elton_name, iucn_name, mass, category, id) %>%
  printtable(caption = "Example IUCN and trait data joined on taxonomic ID. Multilple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.")
```

This results in many more matches (Table 5), as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.  While we have focused on a small example for visual clarity here, the `get_ids()` function in `taxadb` can quickly resolve hundreds of thousands of species names to unique identifiers, thanks to the performance of fast joins in a local MonetDBLite database.

@@

```{r}
multi_match <- filter_name("Zosterops rendovae", provider = "col")
```

```{r echo = FALSE, cache = FALSE}
multi_match %>%
  select(1:5) %>%
  printtable(caption = "Some names may not resolve to an identifier using get\\_ids() because they match to more than one accepted ID. In such cases filter\\_ functions give further detail, as in the example of Zosterops rendovae below which has two accepted ID matches.", escape = FALSE)
```

We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between (Table 6).
This is an example of how `taxadb` seeks to provide users with information from existing authorities and names providers,
rather than make a potentially arbitrary decision.  Because they return `data.frame`s,  `filter_*` functions provide both potential matches.  Note that the simpler `get_` functions (`get_ids()`) consider multiple name as `NA` for the `id`, making them suitable for automated pipelines where manual resolution of duplicates is not an option.

@@

The full taxonomic record in the database can also be directly accessed by `taxa_tbl()`, allowing for whole-database queries that are not possible through the API or web interface of many providers.  For example, we can easily check the coverage of accepted species names in each of the classes of vertabrates within the Catalogue of Life (Table 7):
